Visualizing Independence Using Extended Association Plots

### 2005-08-02 trunk/vcd/paper/JCGS.Rnw
properly included now the "fuzz" for the borders in Friendly shading
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@425 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{center}
\textbf{\LARGE Residual-based Shadings for Visualizing (Conditional) Independence}

@@

\vspace{0.3cm}
\textit{\large Wirtschaftsuniversit\"at Wien, Augasse 2--6, 1090 Wien, Austria}\\[0.14cm]
{\large \textit{E-mail:} \texttt{Firstname.Lastname@wu-wien.ac.at}}
\end{center}

@@

data(UCBAdmissions)
ucb <- aperm(UCBAdmissions)
@

<<auxiliary_functions,echo=FALSE,results=hide>>=
hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}

plot.slice <- function(color.slice)
{
  nc <- ncol(color.slice)
  nl <- nrow(color.slice)
  chroma <- as.numeric(colnames(color.slice))
  luminance <- as.numeric(rownames(color.slice))
  if(attr(color.slice, "type") == "HCL") {
    xlab <- "chroma"
    ylab <- "luminance"
  } else {
    xlab <- "saturation"
    ylab <- "value"
  }
  plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
       xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = deparse(substitute(color.slice)))
  for(i in 1:(nc-1)) {
    rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
  }
  axis(1)
  axis(2)
  box()
}
@

\begin{abstract}

Visualizing independence by association and mosaic plots.

@@

Arthritis data \citep{vcd:Koch+Edwards:1988}

UCB admissions data \citep{vcd:Bickel+Hammel+O'Connell:1975}

@@

@@

We consider a 2-way contingency table with cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
convenience the number of observations is denoted $n = n_{++}$.

Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
measure of discrepancy
between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
Therefore, a rather intuitive idea is to reject the null hypothesis when
there are residuals which are too extreme, i.e., not close enough to
zero. The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test for independence in 2-way tables.
%% reference needed

But this is not the only plausible way of aggregation of the Pearson
residuals. There are many conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda(\{r_{ij}\})$, the sum of squares
is just one of them. Another functional suitable for identifying
the cells which cause the dependence (if any) is the maximum of the absolute
values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
the dependence can easily be identified.

Furthermore, the main reason for using the unconditional limiting distribution
for the $\chi^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
performing permutation tests---either by simulation or by computation of the
permutation distribution---became more and more popular. In particular for
the independence hypothesis (\ref{eq:H0}), using a permutation test is very intuitive
due to the permutation invariance (given row and column sums) of this problem.
By employing this approach the permutation distribution of statistics of
type $\lambda(\{r_{ij}\})$ (also including the $\chi^2$ statistic) can be derived.

Other classical tests for the hypothesis of independence  not fitting in
the Pearson residual-based framework described above include Fisher's exact test,
asymptotic tests of the odds ratio (for $2\times2$ tables) and the Mantel-Haenzel
test.

@@

ONLY SHAPE! not color

The two best known visualization techniques for independence in 2-way tables are
association plots and mosaic plots. Both are suitable to bring out
departures of an observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
The former focuses on the visualization of the Pearson residuals $r_{ij}$ (under
independence) while the latter primarily displays the observed frequencies
$n_{ij}$.

\textbf{Association Plots}
\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The sign of the residual is redundantly
coded by the rectangle's color and its position relative to the baseline.

%% Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
%% home and away goals for all 306 games in the 1995/6 season of the German soccer league
%% Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
%% In particular, it can be seen that there are
%% fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
%% But this does not yet imply the significance of some test statistic.

\textbf{Mosaic Plots}
\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
bar charts where width and height of the bars show the relative
frequencies of the two variables: a mosaic plot simply
consists of a collection of tiles whose
sizes are proportional to the observed cell frequencies
(see Figure \ref{fig:goals-mosaic-classic}).

Sequential horizontal and vertical recursive splits are used to
visualize the frequencies of more than two variables, each new variable
conditional to the previously entered variables. A first extension by
\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under independence. This approach does not only work in 2-way
tables but also in log-linear models fitted to multi-way tables. In this
extension, positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
as in general it is unknown to which significance level $\alpha$ the
values $c_\alpha = 2$ and $4$ correspond for the table being visualized.

@@

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
mysaturation <- as.numeric(colnames(hue23))
value <- as.numeric(rownames(hue23))

@@

## plot 2 slides from HSV space
plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(mysaturation[i], value[-nl], mysaturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-mysaturation[i], value[-nl], -(mysaturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
}
axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
mtext("saturation", side = 3, at = 0, line = 2)
mtext("value", side = 2, at = .75, line = 2)
mtext("value", side = 4, at = .75, line = 2)
lines(c(-1, 1), c(.5, .5))

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
set.seed(1234)
mosaic(art, newpage = FALSE, gp = shading_max, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and extended shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

@@

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9>>=
set.seed(1234)
cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
@
\caption{Conditional association plot for UCB admissions.}
\label{fig:emp-mosaic1}
\end{center}
\end{figure}

>>>>

\begin{center}
\textbf{\LARGE Residual-based Shadings for Visualizing\\[.15cm] (Conditional) Independence}

@@

%% \vspace{0.3cm}
%% \textit{\large Wirtschaftsuniversit\"at Wien, Augasse 2--6, 1090 Wien, Austria}\\[0.14cm]
%% {\large \textit{E-mail:} \texttt{Firstname.Lastname@wu-wien.ac.at}}
\end{center}

@@

data(UCBAdmissions)
ucb <- aperm(UCBAdmissions)

myseed <- 1071
@

<<inference,echo=FALSE,results=hide>>=
if(FALSE) {
  set.seed(myseed)
  art_max <- coindep_test(art, n = 5000)
  set.seed(myseed)
  art_chisq <- coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))

  set.seed(myseed)
  art_gp <- shading_max(art, n = 5000)

  set.seed(myseed)
  ucb_doublemax <- coindep_test(ucb, "Dept", n = 5000)
  set.seed(myseed)
  ucb_maxChisq <- coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2))
  set.seed(myseed)
  ucb_sumChisq <- coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2), aggfun = sum)

  set.seed(myseed)
  ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = 5000)

  save(art_max, art_chisq, art_gp,
       ucb_doublemax, ucb_maxChisq, ucb_sumChisq, ucb_cotab,
       file = "JCGS-inference.rda")
} else {
  load("JCGS-inference.rda")
}
@

<<auxiliary_functions,echo=FALSE,results=hide>>=
stat <- function(x, digits = 3) as.vector(round(x$statistic, digits = digits))
pval <- function(x, digits = 6) {
  oop <- options()
  on.exit(oop)
  options(scipen = 1000)
  as.vector(round(x$p.value, digits = digits))
}

hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}
@

\begin{abstract}
Visualizing independence by association and mosaic plots.

@@

@@

In this section, the basic tools for testing and visualizing 
independence in 2-way tables are briefly reviewed. 
For illustration, a data set about treatment and improvement of
patients with rheumatoid arthritis from \cite{vcd:Koch+Edwards:1988}
is used. The data set is also discussed in \cite{vcd:Friendly:2000}
and the subset of the $59$ female patients from the study is given in
Table~\ref{tab:art}.

@@

To fix notations, we consider a 2-way contingency table with cell
frequencies $\{n_{ij}\}$ for $i = 1, \dots, I$ and $j = 1, \dots, J$
and row and column sums $n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$
respectively.
For convenience the number of observations is denoted $n = n_{++}$.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced at the very beginning
of the chapter about independence in 2-way tables in statistics textbooks
\citep[see e.g.,][]{vcd:Agresti:2002}.

\begin{table}[t!]
\begin{center}
\caption{\label{tab:art} Treatment and improvement among $59$ patients
with rheumatoid arthritis.}
\begin{tabular}{rr|rrr}
                   &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
                   &         & None & Some & Marked \\ \hline
\textbf{Treatment} & Placebo &   19 &    7 &      6 \\
                   & Treated &    6 &    5 &     16
\end{tabular}
\end{center}
\end{table}

But the sum of squares is not the only plausible way of capturing deviations
from $0$ in the Pearson residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda(\{r_{ij}\})$: one which is particularly
suitable for identifying the cells which cause the dependence (if any) is
the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
$\alpha$ \cite[][Ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
the dependence can easily be identified.

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep{vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis (\ref{eq:H0}), using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating from the permutation distribution of test statistics of type
$\lambda(\{r_{ij}\})$.

Other measures of discrepancy (e.g., deviance residuals) could, of course, also be used
instead of the the Pearson residuals $\{r_{ij}\}$. But as the ideas discussed here
extend straightforwardly to that situation, we do not discuss it in detail.

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from Equation~(\ref{eq:ChiSquare})
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistion from Equation~(\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
with a $p$ value of $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
by drawing a sample of size $5000$ from the permutation distribution under independence.

@@

Two well-established visualization techniques for independence in 2-way tables are
mosaic plots and association plots. Both are suitable to bring out
departures of an observed table $\{n_{ij}\}$ from the expected table $\{\hat n_{ij}\}$
in a graphical way. The latter focuses on the visualization of the Pearson residuals
$r_{ij}$ (under independence) while the latter primarily displays the observed frequencies
$n_{ij}$.

\emph{Mosaic plots}
\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
bar charts where width and height of the bars show the relative
frequencies of the two variables: a mosaic plot simply
consists of a collection of tiles whose
sizes are proportional to the observed cell frequencies as shown
in the left panel of Figure~\ref{fig:classic}. A rectangle corresponding
to $100$ percent of the observations is first split horizontally with respect to 
the treatment frequencies and then vertically with respect to the conditional
improvement frequencies. This shows that there have been more placebo than treated 
patients with no improvement and vice versa for marked improvement.
This strategy of splitting with respect to conditional frequencies given all
previous variables can also directly be used for visualizing multi-way tables
\citep[see][for an overview of how to construct and read mosaic displays]{vcd:Hofmann:2003}.

\emph{Association plots}
\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
same interpretation as the mosaic plot: there are more placebo patients with no 
improvement and fewer with marked improvement than expected under independence---vice
versa for the treated patiens.

@@

A first extension by
\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under independence. This approach does not only work in 2-way
tables but also in log-linear models fitted to multi-way tables. In this
extension, positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
as in general it is unknown to which significance level $\alpha$ the
values $c_\alpha = 2$ and $4$ correspond for the table being visualized.

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
saturation <- as.numeric(colnames(hue23))
value <- as.numeric(rownames(hue23))

@@

## plot 2 slides from HSV space
plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
}
axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
mtext("saturation", side = 3, at = 0, line = 2)
mtext("value", side = 2, at = .75, line = 2)
mtext("value", side = 4, at = .75, line = 2)
lines(c(-1, 1), c(.5, .5))

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and extended shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

@@

UCB admissions data \citep{vcd:Bickel+Hammel+O'Connell:1975}
double max: \Sexpr{stat(ucb_doublemax)}, p-value = \Sexpr{pval(ucb_doublemax)}
max Chisq: \Sexpr{stat(ucb_maxChisq)}, p-value = \Sexpr{pval(ucb_maxChisq)}
sum Chisq: \Sexpr{stat(ucb_sumChisq)}, p-value = \Sexpr{pval(ucb_sumChisq)}
simulated from $5000$ repetitions

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9>>=
cotabplot(~ Gender + Admit | Dept, data = ucb, panel = ucb_cotab)
@
\caption{Conditional association plot for UCB admissions.}
\label{fig:emp-mosaic1}
\end{center}
\end{figure}
